HO unification from object language to meta language

### 2024-02-20 paper/main_old.tex
update main

\section{main idea: FO encoding - HO encoding}

\begin{tabular}{c|c}
\begin{minipage}{.5\textwidth}
\begin{minted}{elpi}
kind fo_tm type.
type fo_app list fo_tm -> fo_tm.
type fo_lam (tm -> fo_tm) -> fo_tm.
type fo_uv  nat -> fo_tm.
type fo_c string -> fo_tm.

typeabbrev subst list (option tm).
type equal subst -> tm -> tm -> o.
\end{minted}
\end{minipage}
&
\begin{minipage}{.5\textwidth}
\begin{minted}{elpi}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type uv  nat -> list tm -> tm.
type c string -> tm.

@@

typeabbrev subst list (option assmt).
type equal subst -> tm -> tm -> o.
\end{minted}
\end{minipage}
\end{tabular}

@@

\noindent
\begin{tabular}{c}
\begin{minipage}{.95\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N) T1 :- assigned? N S T, equal S T T1.
equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
equal S (app [uv N|A]) T1 :- assigned? N S F, beta F A T, equal S T T1.
equal S T1 (app [uv N|A]) :- assigned? N S F, beta F A T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N) (uv N).
% beta
equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
% eta
equal S (lam F) T :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
equal S T (lam F) :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
\end{minted}
\end{minipage}
\\\hline
\begin{minipage}{.95\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N A) T1 :- assigned? N S F, beta-abs F A T, equal S T T1.
equal S T1 (uv N A) :- assigned? N S F, beta-abs F A T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
\end{minted}
\end{minipage}

\end{tabular}

>>>>

% \section{main idea: FO encoding - HO encoding}

\section{Lanuages description}

% Description of languages (OL and ML)

In order to reason about unification of the terms of an objet language within a
meta lanauge, we start by formally describing the two languages. We use
meta-programming to accomplish this result. In \cref{code:lang-descr}, we
provide on the left the term representation of the object language. %
% Variables in the OL and ML are represented with integers
In this encoding, the fo\_tm type stands for first order representation of the
object language terms. We model unification variables as integers corresponding
to memory adresses. This memory is represented witht the subst type which is 
list of optional terms. If the cell $i$ is none, then the variable $i$ is not
assigned.

% Why the fo prefix for fo_tm && Uvar in OL have no scope
This encoding is called first order, since variables have no scope. If we take
back the subterm \coqIn{~$\forall$~x,P x} from the forall\_dec instance, we have
\coqIn{P: A ~$\to$~ Prop} and in that subterm, \coqIn{P} has in scope the
variable \coqIn{x}. The translation of this term in the ML is 
\elpiIn{all x\app[uv 1, x]}; \elpiIn{x} is not in the scope of \elpiIn{uv 1}.

% Uvar in ML have scope
To solve this problem, we give more expressiveness to the OL variables, by
giving an alternative term encoding, where unification variables are the
variables of the meta language. These variables have a scope and therefore
can be of higher order.

% The subst mappings
In order to represent variable assignation, we declare the two types subst\_fo
and subst coressponding to the substitution mapping of unification variable
for both the OL and the ML.

\def\eqfo{equal\_fo\xspace}
\def\eqho{equal\_ho\xspace}

% We have two different equal functions
Unification over terms is possible thanks to the equal predicate. There exist
two version of this predicate: \eqfo and \eqho. Their implementation is given in
\cref{code:term-equal}. The first tests if two terms are equal in the object
langauge, that is, given a substitution $\theta$, is it true that the two terms
are equal? Note that no variable instanciation is done: unification is not
performed by the \elpiIn{equal} predicate. Therefore, a unification variabile
$i$ equal to another term $t$ if $t$ is a variable with same index $i$, or if
$i$ is assigend to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover
\eqfo, is able to verify if two terms are equal up to $\eta\beta$.

The \eqho predicate to test if two terms of the

% Compile FO to HO

% Assignement in HO should be decompiled

-------------------------------------------------------------------

\begin{figure*}
  \begin{tabular}{c}
    \begin{minipage}{.35\textwidth}

      \begin{elpicode}
        kind fo_tm type.
        type fo_app list fo_tm -> fo_tm.
        type fo_lam (fo_tm -> fo_tm) -> fo_tm.
        type fo_uv  nat -> fo_tm.
        type fo_c string -> fo_tm.

        typeabbrev subst list (option fo_tm).
        type eq subst -> fo_tm -> fo_tm -> o.
      \end{elpicode}
    \end{minipage}
    \begin{minipage}{.35\textwidth}

      \begin{elpicode}
        kind tm type.
        type app list tm -> tm.
        type lam (tm -> tm) -> tm.
        type uv  nat -> list tm -> tm.
        type c string -> tm.

@@

        typeabbrev subst list (option assmt).
        type eq subst -> tm -> tm -> o.
      \end{elpicode}
    \end{minipage}
  \end{tabular}

  \caption{Language description}
  \label{code:lang-descr}
  \Description[lang-spec]{Language description}
\end{figure*}

% The two equals: equal of OL and ML
% In ML we don't have eta beta -> only unification is allowed on variables

\begin{figure*}
  \begin{tabular}{c}
    \begin{minipage}{.95\textwidth}
      \begin{minted}{elpi}
        % deref
        equal S (uv N) T1 :- assigned? N S T, equal S T T1.
        equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
        equal S (app [uv N|A]) T1 :- assigned? N S F, beta F A T, equal S T T1.
        equal S T1 (app [uv N|A]) :- assigned? N S F, beta F A T, equal S T1 T.
        % congruence
        equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
        equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
        equal _ (c X) (c X).
        equal _ (uv N) (uv N).
        % beta
        equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
        equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
        % eta
        equal S (lam F) T :- not (T = lam _),
          pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
        equal S T (lam F) :- not (T = lam _),
          pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
      \end{minted}
    \end{minipage}
    \\ \hline
    \begin{minipage}{.95\textwidth}
      \begin{minted}{elpi}
        % deref
        equal S (uv N A) T1 :- assigned? N S F, beta-abs F A T, equal S T T1.
        equal S T1 (uv N A) :- assigned? N S F, beta-abs F A T, equal S T1 T.
        % congruence
        equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
        equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
        equal _ (c X) (c X).
        equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
      \end{minted}
    \end{minipage}
  \end{tabular}
  \caption{Term equality}
  \label{code:term-equal}
  \Description[term-equal]{Term equality}
\end{figure*}

@@

\noindent
